---
title: "Tutorial with RIVERS data"
author: "David Cunillera-Montcusí"
date: "2023-07-06"
output: rmdformats::downcute
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  encoding=encoding,
  output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
  rmarkdown::render(input_file,
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,
                      results = "hide",
                      message=FALSE, warning=FALSE)
```

## A tutorial with the RETROMED data to elaborate NATs

Hi! I am David! Basically I wrote everything in Rmarkdown because might
be easy to read and the tutorial with explanations and text next to the
chunks than a script with lots of #. But anyway, this document is to
somehow condensate all the information that I did during the workshop in
Girona un June (12,13,14 of 2023). So, let's begin!

```{r set library  , eval=FALSE}
# Charging packages
library(cooccur) # calculate coocurrance matrices 
library(tidyverse) # managing data.frames and organize and edit them
library(qgraph) # qgraph is a package for generating graphs and containing different types of graph settings
library(network) # a package to manage and create "network" like objects to plot
library(ggnetwork) # plotting networks in a ggplot environment
library(viridis) # colours for colour-sensitive persons 
library(igraph) # classic and mostly used package for network calculation 
```

We charged the packages and all the needed functions. Now we need the
data which basically consists on two things: 1. Traits information for
all the species (a Tachete database that Stephi had) 2. Abundance data
from the RETROMED project that includes 3 pond networks.\
- Albera - Giara - Guils de cerdanya

We transform affinities to work with only presence/absence of traits in
a species. Simply we agreed that might be easier to interpret the
resulting coocurrence network. In this sense, we are just considering
the possibility of trait combinations to exist and we do not consider if
this combinations are "stronger" or weaker".

```{r traits and data , eval=FALSE}
# Traits database 
traits1 <-read.csv("data/tachet.traits.def.csv", header=TRUE, sep=";", na.strings="")
traits2 <-read.csv("data/tachet.traits.def.csv", header=TRUE, sep=";", na.strings="")
# Transform traits to 1 or 0 (losing affiliations)
traits1[,11:ncol(traits1)] <- ifelse(traits1[,11:ncol(traits1)]>=1,1,0)#all afiliations higher than 1 have a 1
traits2[,11:ncol(traits2)] <- ifelse(traits2[,11:ncol(traits2)]>=2,1,0)#only afiliations higher or equal to 2 has a 1.

# Load dataset from retromed
load("data/Rivers/macros_river_list.xlsx")
inv.l # we will use the individual values

head(traits)
head(macros_river_list)
```

We need to filter the traits database with the families that have been found in the
river. 

```{r depuration database , eval=FALSE}
# Check presence
db_indv.l <- macros_river_list %>% filter(taxa%in%traits$Family) # Filter river families from trait database 


head(db_indv.l) #Revisar: falten 82 generes per assignar
save(db_indv.l, file = "ind_match_tachet.RData")
```

Now we have our database ready to go and we can filter by the year
and therefore work for each unit of interest (1995,1996,1997,...2015). The
next step is to built the trait database that we will use to calculate
the coocurrence matrix. This trait database is basically obtaining the
speciesxtrait matrix that we want to use to calculate coocurrences. It
is an easy task and we only need some lines of code to make things match
(specially the lines come to depurate the database that is a bit "messy"
and are just to use them as example). Find below an example with only
Albera network

```{r filter and create the basic dataste , eval=FALSE}

db_indv.l <- db_indv.l %>% filter(year=="1997") # We filter to obtain the ponds of the year 1997

traits_ind <- db_indv.l %>% 
              ungroup() %>% # just in case 
              group_by(taxa) %>% # we group by genus 
              summarise(tot_ab=sum(indv.l)) %>% # we sum the abundance of genus in the whole network
              left_join(traits, by=c("taxa"="Family"),multiple ="all") %>% # join traits with the genus of the traits database
              group_by(taxa) %>% # regroup by genus
              filter(n()<2) %>% # filter those species that are listed 2 times (by mistake), check which are these in case that there are some. It means mistakes.
              select(c(1,12:ncol(.))) %>% # select the columns with only traits
              na.omit() %>% # NA elimination
              tibble::column_to_rownames("taxa") # we attach genus as rownames

head(traits_ind)
load("traits_ind.RData")
save(traits_ind, file = "traits_ind.RData")
```

Everything is ready! This is the type of table that we need to have for
running our trait coocurrence matrix. We have the *genus* in each row
and the *traits* (in 1/0) in each column. This is the "key" of all the
dynamics of the exercise. So, this table is the "core" of all the
exercise, because this table will be the result of considering more or
less ponds of the network. As a consequence, each change in this
database will have a consequence in the network and the values that we
will obtain later. So, it is in building this table that we are actually
"doing" the work (adding more or less sites, adding sites in a different
way, etcetera)

So let's do the coocurrence analysis!

```{r coocurrence , eval=FALSE}
#run analysis (takes a bit)
cooccur.species_ccr <- cooccur(mat = t(traits_ind), type = "spp_site", thresh = T, spp_names = TRUE)
# this script comes from the supplementary material of the "original" NAT's paper (DOI: 10.1002/eap.2010)
# Note that we transpose the matrix (t())

head(cooccur.species_ccr)
```

As we can see there is a lot of results but basically we have columns
indicating each "sp" (AKA=traits) and their combinations and the
"prob_coocur" column which is indicating the probability of these two
"species" being found in the same site (in other words, coocurring).
Remember that in our case sp=traits site=genus. In summary, it is giving
us the probability of a combination of traits (Sp1 and Sp2) is found in
the community that we considered and therefore, how often do we observe
this combination of traits in our community species.

This logic at the end is the "ecological" meaning that we give to the
links and the networks. If all the communities are very similar among
them the trait combinations that will have high probabilities of being
together will be the same all the time and the network will "look the
same". On the other hand if they differ, the network will start to
change and we will detect that alteration.

So let's built a network out of the information that we have. Actually,
the network is already built because we already have "who is connected
to who" and the strength of this connection. Sp1 and Sp2 are indicating
who is connected to who and the "prob_coocur" is the strength of this
link. To built the network we just need to translate this information
into a matrix of traits x traits (or in network/"igraph" language, an
adjacency matrix).

```{r network matrix and graph , eval=FALSE}
# Generate an adjacency table from the coocurrence matrix 
## Create a blank table with only 0 
adj_table <- matrix(nrow =cooccur.species_ccr$species,ncol = cooccur.species_ccr$species,data = 0)

for (row_ID in 1:length(unique(cooccur.species_ccr$results$sp1))) {# do the following for each spp
# Summarise the strenght of the link for each genus by calulating the mean probability
link_weight <- cooccur.species_ccr$results %>% group_by(sp1,sp2) %>% summarise(mean_prob=mean(prob_cooccur))
# filter the results from the coocurrence table for a species. You isolate all the species with who the 
## targeted species is coocuring
spp_to <- cooccur.species_ccr$results %>%filter(sp1==unique(cooccur.species_ccr$results$sp1)[row_ID])
# You add the values into the "adj.table" that was empty and now is being filled
adj_table[unique(spp_to$sp1),spp_to$sp2] <- spp_to$prob_cooccur
}
# We transpose the results to make the matrix symetrical (same from & to links)
adj_table[lower.tri(adj_table)] <- t(adj_table)[lower.tri(adj_table)]

# We built a graph from the adjacency table
g <- igraph::graph.adjacency(adj_table,mode="undirected",weighted = TRUE)
```

Well done! It is already done! With this you just generated a graph (g)
that contains de information of your coocurrence netowrk of traits! So,
you now can calculate whatever value from this graph that informs you
about how this network is structured. There is plenty of metrics and
variables that you can calculate BUT you must be aware of what they mean
and more importantly "on what" you are calculating these metrics. In
general there are "global" metrics (e.g., diameter, edge density) or
"per node" metrics (e.g. degree, closeness, betweenness). For global
metrics you will obtain 1 value for the whole network while for "per
node" metrics you will get a value for each one of the nodes of your
network (in our case for each trait). ALSO, be aware of the "scale" that
each metric is considering! Degree counts de number of neighbors that a
node has while closeness average the path lengths to all network
nodes... this implies that different metrics can capture connectivity at
different levels of the network!

Last warning! In general we use "igraph" package (kind of the trendy
one) but there is also "sna" that can be quite useful for some things.
HOWEVER, both packages have functions with the same names!!! This can
generate conflicts and we must be careful with it (I tend to write
igraph::function or sna::function to always be sure) ;)

```{r some network metrics , eval=FALSE}
edge_dens_man <- length(igraph::E(g))/(((ncol(traits_ind)*(ncol(traits_ind)-1)))/2)
edge_dens <- igraph::edge_density(g)
mean_grStre <- mean(igraph::graph.strength(g))
sd_grStre <- sd(igraph::graph.strength(g))
medi_grStre <- median(igraph::graph.strength(g))
min_grStre <- min(igraph::graph.strength(g))
max_grStre <- max(igraph::graph.strength(g))
```

Finally, what is always more nice to do with networks is to plot them!
However, this can be a bit messy as there are several packages and
things to use... it is a bit as doing a basic plot, there is 1934320
ways to do it and 230493204923 packages to achieve it. I am quite use to
work with "tidyverse" and therefore the "ggplot" kind of approach. So,
my recommendation is based on that but I found that is the most easy
(for me!).

In our NAT however, the representation of the network is a bit meh...
because we do not have a "distance" or an x,y coordinates that can be
used as a "layout". For this, we use "The Fruchterman-Reingold layout
algorithm" that it basically considers the weight of each vertice (AKA
node) to place it in the space (kind of a "distance" measure).

```{r Fruchterman-Reingold , eval=FALSE }
# Layout plot (distance between nodes based on the links weight)
l <- igraph::layout_with_fr(g,start.temp = 1500)#start.temp is what allows more or less "movement" along the axis. High values make more "sparse" graphs but also they are "far" from reality. Smaller values make more "dense" graphs that provide less visual information. Balanxce between both levels must be achieved. 
colnames(l) <- c("x","y")
```

Once we have the "layout" we can built our ggplot. Basically I use the
setting provided by the network and ggnetwork packages that allow to
draw the newtork in a ggplot-like enviornment. To do this, we first need
to "destroy" our previous graph into a matrix than can be understood as
a "network" object. Then we will add the parameters that we want to use
to plot and color the graph. Depending if we want to add these
parameters to the nodes (vertices) or to the links (edges) we will need
to add %v% or %e% correspondingly. This will make much more easy the
setting of colour/fill/size/alpha and other aesthetic parameters.

```{r plot  , eval=FALSE}
library(network);library(ggnetwork);library(viridis) # Packages

n<- network(as.matrix(igraph::as_adjacency_matrix(g)), directed=F) # Newtork creation
# Setting the parameters that we want to use. You can add as many parameters as you wish. 
n %v% "Mean_Prob" <- igraph::eigen_centrality(g)$vector # Weighted degree
n %v% "FreqTRAIT" <-apply(ifelse(traits_ind>0,1,0),2,sum) # Trait frequency
n %e% "Edg_values" <-igraph::E(g)$weight # Edge weights (coocurrence probability)

# The plot per se
ggplot(n, 
       layout=as.matrix(l),aes(x = x, y = y, xend = xend, yend = yend))+ # We need to set the layout that we created. Remember that this is just to have an x and y coordinates. 
  geom_edges(aes(color =Edg_values, size=Edg_values),arrow=arrow(angle = 20),curvature = 0.15)+ # Information and aesthetics of the edges
  geom_nodes(aes(size=FreqTRAIT, fill=Mean_Prob),color="black" ,shape=21)+ # Information and aesthetics of the edges
  # From here onwards everythign is like a normal ggplot... 
  scale_color_viridis(direction = -1,option = "D")+
  scale_fill_viridis(direction = -1,option = "A")+labs(title="")+
  theme_classic()+
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background=element_blank())
```
